---
title: "Assignment 2: Coding Basics"
author: "Devin Domeyer"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. I am telling R to create a sequence from 1 to 100, increasing by 4.
sequence <- seq(1, 100, 4)
sequence

#2. I am telling R to find the mean and median of the sequence
mean <- mean(sequence)
median <- median(sequence)

#3. The output of this conditional statement will either be true or false.
mean > median

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
# The vector "name" is numeric
name <- c("Todd", "Sally", "Ron", "Sandra")
# The vector "score" is character
score <- c(42, 85, 63, 94)
# The vector "pass_fail" is logical
pass_fail <- c(FALSE, TRUE, TRUE, TRUE)

# Now I combine the new vectors into a single dataframe using cbind
test_stats <- cbind(name, score, pass_fail)

# Now I change the column names
colnames(test_stats) <- c("Student_Name", "Test_Score", "Pass/Fail")

test_stats

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: 

A data frame is a more general form of data structure than a data matrix. In a matrix all the columns have the same mode and the same length. In a data frame, you can have different modes. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

pass_test <- function(x) {
  ifelse(x>50, TRUE, FALSE)

}

pass_test(score)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 

ifelse worked becuase I put in my logical expression first "if x is greater than 50", then if that expression matched the vector value it was "TRUE" and if it didn't it was "FALSE". It would have worked if I separated out the "if" and "else" function elements but I wanted to keep my code clean.
